---
title: "STAT-6494 Advanced Statistical Computing with R"
subtitle: "Homework 1"
author: Wenjie Wang
date: "`r format(Sys.time(), '%d %B %Y')`"
header-includes:
    - \usepackage{booktabs}
    - \usepackage{bm}
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
    number_sections: no
    toc: no
  fontsize: 11pt
  classoption: letter
  documentclass: article

---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
## function ipak: install and load multiple R packages.
## check to see if packages are installed.
## install them if they are not, then attach them to the search path.
ipak <- function (pkg) {
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg))
        install.packages(new.pkg, dependencies = TRUE)
    sapply(pkg, require, character.only = TRUE, quietly = TRUE)
    invisible(NULL)
}
## install packages needed
needpack <- c("microbenchmark")
ipak(needpack)
## knitr::opts_chunk$set(dev = 'pdf')

## install package "comorbidities"
if (! require(comorbidities)) {
    download.file(paste("https://cran.r-project.org/src/contrib/Archive/",
                        "comorbidities/comorbidities_1.0.tar.gz", sep = ""),
                  destfile = "comorbidities_1.0.tar.gz", quiet = TRUE)
    install.packages("comorbidities_1.0.tar.gz", repos = NULL)
}

```


# Exercise: Comorbidity Measures

Comorbidity measures are important in health-care research. They can be obtained
from hospital discharge records using the diagnosis coding of ICD-9-CM
(International Classification of Diseases, Ninth Edition, Clinical Modifications).
An influential, widely cited method was proposed by Elixhauser et
al. (Comorbidity Measures for Use with Administrative Data. Medical Care,
1998;36:8–27). The method has been refined by the Healthcare Cost and
Utilization Project [HCUP](http://www.hcup-us.ahrq.gov/overview.jsp).
Their [implementation
](http://www.hcup-us.ahrq.gov/toolssoftware/comorbidity/comorbidity.jsp)
is in SAS. Construction of the variables is summarized in a PDF table under the
section "Variables Created". For simplicity, let us ignore the DRG code for this
assignment. For real usage, one would need to consider the DRG code.  An R
implementation is available in package
[comorbidities](cran.r-project.org/package=comorbidities).
We will focus on the function ahrq in this package. The input of the function is
a data frame where rows are patients and columns are ICD-9-CM codes. Each
patient does not necessarily have all columns filled; some ICD-9-CM codes can be
NA or empty. The output of the functions is a list, and we are interested in the
data frame where rows are patients and columns are comorbidity measures. This
function can be made much more efficient by vectorizing.  Please write a
function `cmbd`. Its input is the same as ahrq and its output is a data frame with
only first two comorbidity measures: CHF and VALVE (again, for simplicity). We
have a sample input of 100 patients in csv format for your testing. The function
will be graded with another sample input randomly selected from my real data.
Comment: For real implementation, We had to clean the data: 1) trim the leading
and trailing spaces; 2) add zero to the end for codes with length smaller than
5; 3) consider DRG code; and 4) automate the generation of functions.


# Implementation: Function `cmbd`

The function body of `cmbd` is shown as follows:

```{r fun, echo = TRUE, comments = NA}
cmbd <- function (data) {
    mat <- as.matrix(data)
    numMat <- gsub("^[Vv]", "32", mat)
    numMat <- gsub("^[Ee]", "15", numMat)
    numMat <- as.numeric(numMat)
    dim(numMat) <- dim(mat)
    funList <- codeFun()
    outDat <- sapply(seq_along(funList),
                     function (a, numMat) {
                         funList[[a]](numMat)
                     }, numMat = numMat)
    colnames(outDat) <- c("CHF", "VALVE")
    data.frame(outDat)
}

codeFun <- function () {
    ## internal function for CHF measures
    chf <- function (numMat) {
        indVec1 <- numMat %in% c(39891, 40201, 40211, 40291, 40401,
                                 40403, 40411, 40413, 40491, 40493)
        indVec2 <- numMat >= 42800 & numMat <= 42890
        indMat <- indVec1 | indVec2
        dim(indMat) <- dim(numMat)
        as.integer(rowSums(indMat, na.rm = TRUE) > 0)
    }
    ## internal function for VALVE measures
    valve <- function (numMat) {
        indVec1 <- numMat >= 9320 & numMat <= 9324
        indVec2 <- numMat >= 39400 & numMat <= 39710
        indVec3 <- numMat >= 42400 & numMat <= 42499
        indVec4 <- numMat >= 74630 & numMat <= 74660
        indVec5 <- numMat %in% c(324220, 324330, 39790)
        indMat <- indVec1 | indVec2 | indVec3 | indVec4 | indVec5 
        dim(indMat) <- dim(numMat)
        as.integer(rowSums(indMat, na.rm = TRUE) > 0)
    }
    list(chf = chf, valve = valve)
}
```


# Result of Sample Dataset

By using the function `cmbd` defined in last section, the diagnosed result
for each patient by ICD9 code in sample dataset is shown as follows.

```{r res, echo=TRUE, warning=FALSE}
sampleDat <- read.csv("icd9sample.csv")
(myRes <- cmbd(sampleDat))
## compare the results with the one from package comorbidities
library(comorbidities)
pkgRes <- (ahrq(sampleDat)[[2]])[, 1:2]
all.equal(pkgRes, myRes) # match if TRUE
```


# Performance

Although function `cmbd` only generate the diagnosis result for CHF and VALVE,
from the comparison of computing performance by package `microbenchmark`, we
may find that the computing speed of `cmbd` is much faster than
function `ahrq` in pacakge `comorbidities`.


```{r perf, cache=TRUE, warning=FALSE}
library(microbenchmark)
microbenchmark(ahrq(sampleDat), cmbd(sampleDat), times = 5)
```
