---
title: "STAT-6494 Advanced Statistical Computing with R -- HW 4"
author: Wenjie Wang
date: "`r format(Sys.time(), '%d %B %Y')`"
header-includes:
    - \usepackage{booktabs}
    - \usepackage{bm}
bibliography: hw4.bib
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
    number_sections: yes
    toc: no
  fontsize: 11pt
  classoption: letter
  documentclass: article

---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
## function ipak: install and load multiple R packages.
## check to see if packages are installed.
## install them if they are not, then attach them to the search path.
ipak <- function (pkg) {
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg))
        install.packages(new.pkg, dependencies = TRUE)
    sapply(pkg, require, character.only = TRUE, quietly = TRUE)
    invisible(NULL)
}

## install packages needed
needpack <- c("ggplot2", "copula", "kfigr", "knitr", "cobs")
ipak(needpack)
knitr::opts_chunk$set(dev = 'pdf')
knitr::opts_chunk$set(comment = NA)
## load data and clean

## source code

## remedy for set-up of cross-reference index of figures and tables
noprint <- function (x) {
  invisible(x)
}
## usage: `r noprint(figr("chunkname", TRUE, type = "Figure"))`

```


# Exercise:  Spearman's rho for $t$-copula

Consider a $t$-copula with two parameters: degrees of freedom df and dispersion
$r$. For any given df, Spearman's rho is a function of $\rho$, but this function
has no closed-form. One way to approximate this function is through sample
Spearman's $\rho$ on a grid of $r$ values in $[-1, 1]$. A spline approximation
can be fitted to the pairs of $\hat{\rho}$ and $r$
[@kojadinovic2010comp, Appendix].

1. Create a grid of $(\hat{\rho},r)$ pairs for a grid of $r$ values with sample
size $n=1000$, with df 5.

2. Find a spline function that approximates the true function $\rho(r)$.

3. Repeat with df 10, and overlay the two functions.

4. It is known that $\rho(-1)=-1$, $\rho(0)=0$, and $\rho(1)=1$.  Can you
incorporate these in the fitting?


## Part 1.

With the help of **R** package **copula**,
we may easily get random numbers from bivariate $t$-copula for given
degree of freedom, df, and dispersion $r$. Then the sample Spearman
correlation coefficient $\hat{\rho}$ is computed for each sample.
Function `oneRho` and `rRho` are implemented to get one $\rho$
at given $r$ and the mean $\rho$ of replicates
at each grid point of $r$ values, respectively. 

```{r fun}
library(copula)
oneRho <- function (r, df, nSample) {
    if (r == 0L) return(0) 
    obj <- tCopula(r, dim = 2, dispstr = "ex", df = df, df.fixed = TRUE)
    sampMat <- rCopula(nSample, obj)
    cor(sampMat[, 1], sampMat[, 2], method = "spearman")
}
rRho <- function (rVec, df = 5, nSample = 1000) {
    rhoVec <- sapply(rVec, oneRho, df = df, nSample = nSample)
    data.frame(r = rVec, rho = rhoVec)
}
```

In part 1, the degree of freedom is fixed to be 5.
The sample size is fixed as 1,000.
We take a grid of $r$ values in $[-1, 1]$ by step $0.01$.


```{r df5, cache = TRUE}
set.seed(1216)
rVec <- seq(- 1, 1, by = 0.01)
resMat1 <- rRho(rVec, df = 5)
```

We may plot the mean $\rho$ over the grid of $r$ as follows:

```{r plotDf5, cache = TRUE, fig.cap = "Sample Spearman's rho (blue solid curve) and reference line with slope 1 and intercept 0 (black dotted line).", fig.width = 6, fig.height = 5}
library(ggplot2)
resDat1 <- data.frame(resMat1, df = 5)
(p1 <- ggplot(resDat1, aes(x = r, y = rho)) +
     geom_point(color = "royalblue", size = 0.8, alpha = 0.8) +
     geom_abline(slope = 1, color = "black", linetype = 3) +
     ylab("Sample Spearman's rho") + theme_bw())
```


## Part 2.

From `r figr("plotDf5", TRUE, type = "Figure")`, We may observe that
the mean Spearman's rho curve looks quite linear. Therefore, penalized
smoothing cubic spline is probably a good choice since it penalizes the
second derivatives. The tuning parameter is left to be selected by GCV.


```{r spline1, cache = TRUE, fig.cap = "Fitted Spearman's rho curve by penalized spline.",fig.height = 6, fig.width = 6, cache = TRUE}
(smoothFit1 <- with(resDat1, smooth.spline(x = r, y = rho)))
plot(rho ~ r, data = resDat1, col = "gray",
     xlab = expression(r), ylab = expression(rho))
lines(smoothFit1, col = "blue")
abline(a = 0, b = 1, col = 1, lty = 3)
legend("bottomright", col = c("blue", "gray", "black"),
       legend = c("Fitted spline", "Mean Spearman rho",
                  expression("Reference line"~y==x)),
       lty = c(1, NA, 3), pch = c(NA, 1, NA), cex = 0.8)
```

The fitted spline function is plotted in
`r figr("spline1", TRUE, type = "Figure")`, from which we may find that
the penalized smoothing cubic spline fits the mean Spearman's rho quite well. 


## Part 3.

When df is specified as 10, the similar results follow.

```{r df10, cache = TRUE}
set.seed(1216)
resMat2 <- rRho(rVec, df = 10)
```

```{r plotDf10, cache = TRUE, fig.cap = "Sample Spearman's rho (red solid curve) and reference line with slope 1 and intercept 0 (black dotted line).", fig.width = 6, fig.height = 5}
resDat2 <- data.frame(resMat2, df = 10)
(p2 <- ggplot(resDat2, aes(x = r, y = rho)) +
     geom_point(color = "red", alpha = 0.8, cex = 0.8) +
     geom_abline(slope = 1, color = "black", linetype = 3) +
     ylab("Sample Spearman's rho") + theme_bw())
```


```{r spline2, cache = TRUE, fig.cap = "Fitted Spearman's rho curve by penalized spline.",fig.height = 6, fig.width = 6, cache = TRUE}
(smoothFit2 <- with(resDat2, smooth.spline(x = r, y = rho)))
plot(rho ~ r, data = resDat2, col = "gray",
     xlab = expression(r), ylab = expression(rho))
lines(smoothFit2, col = "red")
abline(a = 0, b = 1, col = 1, lty = 3)
legend("bottomright", col = c("red", "gray", "black"),
       legend = c("Fitted Spearman's rho", "Sample Spearman rho",
                  expression("Reference line"~y==x)),
       lty = c(1, NA, 3), pch = c(NA, 1, NA), cex = 0.8)
```

We may further overlay two fitted spline functions as follows:

```{r overlay, cache = FALSE, fig.cap = "Overlaid spline functions fitted for t-copula with df 5 and 10.", fig.height = 5, fig.width = 6}
ggDat <- data.frame(r = c(smoothFit1$x, smoothFit2$x),
                    rho = c(smoothFit1$y, smoothFit2$y),
                    df = gl(2, length(rVec), labels = c("5", "10")))
(p3 <- ggplot(ggDat, aes(x = r, y = rho, color = df, linetype = df)) +
     geom_line() + geom_abline(slope = 1, color = 1, linetype = 3) +
     ylab("Fitted Spearman's rho") + theme_bw())
```

The fitted spline function for df$=5$ has slightly larger deviation to
the reference line $y=x$ than df$=10$.


## Part 4.

For simplicity, we fix df$=5$ here.
It is known that the true function of Spearman's rho is symmetric about
zero. Therefore, we may consider fit the smoothing spline ranging from only
0 to 1. The fits over $-1$ to 0 can be obtained by symmetry.

### Weighted smoothing spline

Notice that if $r$ is specified to be 1, the Spearman's rho is
necessarily 1. The function `rRho` produces Spearman's rho as 0, if $r$ is 0.
Therefore, we consider weighted smoothing spline to constrain fitted curve
to be close to $(0, 0)$ and $(1, 1)$. The $(-1, -1)$ will be approached by symmetry.
The weight on data points $(0, 0)$ and $(1, 1)$ is set to be $1,000$, which is much
larger than the weight on other data points.

```{r wSmooth, cache = TRUE, fig.cap = "Fitted Spearman's rho by weighted smoothing spline and symmetry."}
## generate mirror part from fits over 0 and 1
mirFun <- function (fitObj) {
    if (class(fitObj) == "smooth.spline")
        mirDat <- dat <- with(fitObj, cbind(x, y))
    else if (class(fitObj) == "cobs")
        mirDat <- dat <- with(fitObj, cbind(x, fitted))
    else stop("not applicable")
    mirDat <- - mirDat
    outMat <- rbind(tail(mirDat, -1) , dat)
    out <- outMat[order(outMat[, "x"]), ]
    row.names(out) <- NULL
    colnames(out) <- c("r", "rho")
    data.frame(out)
}
rVec <- seq(0, 1, by = 0.01)
mirDat <- rRho(rVec)
(conFit <- with(mirDat, smooth.spline(x = r, y = rho,
                                      w = c(1e3, rep(1, length(r) - 2), 1e3))))
resDat <- mirFun(conFit)
(p4 <- ggplot(resDat, aes(x = r, y = rho)) + geom_line(color = "royalblue") +
     geom_abline(slope = 1, color = 1, linetype = 3, alpha = 0.8) +
     ylab("Fitted Spearman's rho") + xlab("r") + theme_bw())
```

If we check the fitted Spearman's rho at 0 or 1 ($-1$), we will find that
the difference between true value and fitted value is negligible. 

\newpage

### Qualitatively constrained smoothing splines

Another approach that we may consider is
qualitatively constrained smoothing splines [@ng2007fast].
The implementation is available in package **cobs**.
The smoothing parameter is set to be chosen by SIC.

```{r cobs, cache = TRUE, message = FALSE, fig.cap = "Fitted Spearman's rho by qualitatively constrained smoothing spline and symmetry."}
library(cobs)
con <- matrix(c(0, 0, 0, 0, 1, 1), ncol = 3, byrow = TRUE)
cobsFit <- with(mirDat, cobs(x = r, y = rho, constraint = "convex",
                             pointwise = con, lambda = - 1))
resDat <- mirFun(cobsFit)
(p5 <- ggplot(resDat, aes(x = r, y = rho)) + geom_line(color = "red") +
     geom_abline(slope = 1, color = 1, linetype = 3, alpha = 0.8) +
     ylab("Fitted Spearman's rho") + xlab("r") + theme_bw())
```

If we have a close check on the fitted Spearman's rho on 0 or 1 ($-1$),
we may find that the fitted values are not exactly constrained to be the
true values. However, the difference is negligible. 

\newpage

# Reference
