---
title: "STAT-6494 Advanced Statistical Computing with R"
subtitle: "Homework 7"
author: Wenjie Wang
date: "`r format(Sys.time(), '%d %B %Y')`"
header-includes:
    - \usepackage{booktabs}
    - \usepackage{bm}
bibliography: hw7.bib
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
    number_sections: yes
    toc: no
  fontsize: 11pt
  classoption: letter
  documentclass: article

---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
## function ipak: install and load multiple R packages.
## check to see if packages are installed.
## install them if they are not, then attach them to the search path.
ipak <- function (pkg) {
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg))
        install.packages(new.pkg, dependencies = TRUE)
    sapply(pkg, require, character.only = TRUE, quietly = TRUE)
    invisible(NULL)
}

## install packages needed
needpack <- c("knitr")
ipak(needpack)
knitr::opts_chunk$set(comment = ">")
## load data and clean

## source code

```


# Exercise: Estimating the Median with Bootstrap

@efron1979 had an example of estimating the median Section 3. Suppose
that $X_1,\ldots,X_n$, $n=2m-1$ from a random sample from a distribution
$F$. Let $\theta(F)$ be the median of $F$ and $t(\bm{X})=X_{(m)}$ be
the sample median. Having observed $\bm{X}=\bm{x}$ with $n=13$ from
$N(0,1)$, use Monte Carlo bootstrap method to approximate the sampling
distribution of
$$
R(\bm{X}, F) = \frac{|t(\bm{X}) - \theta(F)|}{\sigma(F)},
$$
the absolute error of the sample median relative to the population
standard deviation.


# Reproduction of Example Simulation Study


The following is the function we revised from the course notes and the
corresponding simulation results generated.


```{r fun}
median.boot <- function(x, nboot) {
    tx <- median(x)
    sx <- sd(x) / sqrt(nx <- length(x))
    do1rep <- function(x) {
        x.b <- sample(x, nx, replace = TRUE)
        abs(median(x.b) - tx) / sx
    }
    r.sample <- replicate(nboot, do1rep(x))
    c(mean(r.sample), sd(r.sample))
}
set.seed(421)
x <- rnorm(13)
median.boot(x, 50)
sim <- replicate(10, median.boot(rnorm(13), 100))
resDat <- data.frame(t(sim))
colnames(resDat) <- c("Ave.", "S.D.")
resDat
## average of bootstrap estiamtes and sd estimates
colMeans(resDat)
## empirical sd of bootstrap estimates and sd estimates
apply(resDat, 2, sd)
```


The results shown in the first column resemble the second column in Table 1 on
page 8 in @efron1979. Therefore, the similar results to the example simulation
study is produced successfully.  Note that the only revise we did on the
function `median.boot` from course notes is $\sigma(F)$, the so-called
"population standard deviation", which is introduced to help the quantity of
interest, $R(\bm{X},F)$ be more numerically stable.


# Reference
