---
title: "STAT-6494 Advanced Statistical Computing with R"
subtitle: "Homework 6"
author: Wenjie Wang
date: "`r format(Sys.time(), '%d %B %Y')`"
header-includes:
    - \usepackage{booktabs}
    - \usepackage{bm}
bibliography: hw6.bib
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
    number_sections: yes
    toc: no
  fontsize: 11pt
  classoption: letter
  documentclass: article

---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
## function ipak: install and load multiple R packages.
## check to see if packages are installed.
## install them if they are not, then attach them to the search path.
ipak <- function (pkg) {
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg))
        install.packages(new.pkg, repos = "https://cloud.r-project.org")
    sapply(pkg, require, character.only = TRUE, quietly = TRUE)
    invisible(NULL)
}

## install packages needed
needpack <- c("knitr")
ipak(needpack)
knitr::opts_chunk$set(comment = ">")
## load data and clean

## source code

```


# Exercise: A Poisson-HMM for Earthquake Data

Consider a $m$-state HMM where the observed data follows a Poisson distribution
with mean $\lambda_i$ for state $i$, $i=1,\ldots,m$, respectively.  This model
have three sets of parameter: initial distribution $\delta$, transition
probability matrix $\Gamma$, and Poisson means
$\bm{\lambda}=(\lambda_1, \ldots, \lambda_m)$.
Suppose the observed data is a vector $x$. Write a function `poisson.hmm.em`
that implements the EM algorithm for this model with these argument:

* `x`: the observed data;
* `m`: the number of states;
* `lambda`: initial value of $\lambda$;
* `Gamma`: initial value of $\Gamma$;
* `delta`: initial value of $\delta$;
* `control`: a named list similar to `glm.control` that specifies the tolerance,
maximum number of iteration, and whether or not trace the iteration.

Apply the function to the count of major earthquakes (magnitude 7 or above) in
the world from 1900 to 2015 with a 2-state HMM and a 3-state HMM.

```{r exer}
## frequency of major earthquake in the world from 1900 to 2015
## raw data accessed at http://earthquake.usgs.gov/earthquakes/search/
qk7freq <- c(3, 2, 4, 1, 2, 5, 8, 3, 2, 5, 5, 7, 3, 4, 6, 4, 8, 5, 12, 8, 7, 9,
             7, 12, 9, 12, 13, 11, 16, 15, 9, 19, 9, 8, 12, 14, 11, 9, 21, 14,
             7, 13, 11, 18, 13, 5, 10, 13, 11, 9, 13, 11, 7, 9, 6, 10, 8, 21, 8,
             8, 13, 12, 10, 17, 12, 18, 9, 11, 22, 14, 17, 20, 16, 9, 11, 13,
             14, 10, 12, 8, 6, 10, 7, 14, 14, 15, 11, 13, 11, 9, 18, 17, 13, 12,
             13, 20, 15, 16, 12, 18, 15, 16, 13, 15, 16, 11, 11, 18, 12, 17, 24,
             20, 16, 19, 12, 19)

## forward/backward probability for Poisson-HMM from
## Zucchini and MacDonald (2009):
## Hidden Markov Models for Time Series: An Introduction with R.
pois.HMM.lalphabeta <- function(x, m, lambda, gamma, delta = NULL) {
    if (is.null(delta))
        delta <- solve(t(diag(m) - gamma + 1), rep(1, m))
    n <- length(x)
    lalpha <- lbeta <- matrix(NA, m, n)
    allprobs <- outer(x, lambda, dpois)
    foo <- delta * allprobs[1, ]
    sumfoo <- sum(foo)
    lscale <- log(sumfoo)
    foo <- foo/sumfoo
    lalpha[, 1] <- log(foo) + lscale
    for (i in 2:n) {
        foo <- foo %*% gamma * allprobs[i, ]
        sumfoo <- sum(foo)
        lscale <- lscale + log(sumfoo)
        foo <- foo / sumfoo
        lalpha[, i] <- log(foo) + lscale
    }
    lbeta[, n] <- rep(0, m)
    foo <- rep(1/m, m)
    lscale <- log(m)
    for (i in (n - 1):1) {
        foo <- gamma %*% (allprobs[i + 1, ] * foo)
        lbeta[, i] <- log(foo) + lscale
        sumfoo <- sum(foo)
        foo <- foo/sumfoo
        lscale <- lscale + log(sumfoo)
    }
    list(la = lalpha, lb = lbeta)
}

```


# Implementation

With the notation in the course note and @zucchini2009hidden, the
complete-data-log-likelihood (CDLL) is
$$
\log\left(\mathrm{Pr}(\bm{x}^{(T)}, \bm{c}^{(T)})\right) =
\sum_{j=1}^m \mu_j(1)\log\delta_j + \sum_{j=1}^m \sum_{k=1}^m
\left(\sum_{t=2}^Tv_{jk}(t)\right)\log\gamma_{jk} +
\sum_{j=1}^m \sum_{t=1}^T\mu_j(t)\log p_j(x_t).
$$

## E-step

$$
\hat{\mu}_j(t)=\mathrm{Pr}(C_t=j|\bm{x}^{(T)}) = \alpha_t(j)\beta_t(j)/L_T,
$$
and
$$
\hat{v}_{jk}(t)=\mathrm{Pr}(C_{t-1}=j, C_t=k|\bm{x}^{(T)}) =
\alpha_{t-1}(j)\gamma_{jk}p_k(x_t)\beta_t(k)/L_T,
$$
where $L_T = \bm{\alpha}_t \bm{\beta}_t^{'}$.

## M-step

We may find that the CDLL can be splited into three seperate maximizations,
which lead to the closed form solutions as follows:

1. For $\bm{\delta}$:
$$
\hat{\delta}_j= \arg\max \sum_{j=1}^m\hat{\mu}_j(1) \log\delta_j =
\hat{\mu}_j(1)/\sum_{j=1}^m\hat{\mu}_j(1) = \hat{\mu}_j(1).
$$

2. For $\bm{\Gamma}$:
$$
\hat{\gamma}_{jk} = \arg\max \sum_{j=1}^m \sum_{k=1}^m
\left(\sum_{t=2}^T\hat{v}_{jk}(t)\right) \log\gamma_{jk} =
f_{jk} / \sum_{k=1}^m f_{jk},
$$
where $f_{jk} = \sum_{t=2}^T \hat{v}_{jk}(t)$.

3. For $\bm{\lambda}$:
$$
\hat{\lambda}_j = \arg\max\sum_{j=1}^m \sum_{t=1}^T \hat{\mu}_j(t)\log p_{j}(x_t) =
\sum_{t=1}^T \hat{\mu}_j(t)x_t / \sum_{t=1}^T\hat{\mu}_j(t).
$$

## Function `poisson.hmm.em`


```{r fun}
poisson.hmm.em <- function (x, m, lambda, Gamma, delta,
                            control = list(), ...) {
    control <- do.call(stats::glm.control, control)
    if (missing(lambda)) lambda <- seq(mean(x), by = 1, length.out = m)
    if (missing(Gamma)) Gamma <- matrix(1 / m, m, m)
    if (missing(delta)) delta <- rep(1 / m, m)
    n <- length(x)

    if (control$trace) {
        cat("Initial values:\n", sep = "")
        print(list(lambda = lambda, Gamma = Gamma, delta = delta))
        cat(rep("-", 70), "\n", sep = "")
    }

    for (iter in seq_len(control$maxit)) {
        est <- do1step(lambda, Gamma, delta, m, n, x)
        lambda_new <- est$lambda
        Gamma_new <- est$Gamma
        delta_new <- est$delta
        tol <- sum((lambda - lambda_new) ^ 2) / sum(lambda ^ 2) +
            sum((Gamma - Gamma_new) ^ 2) / sum(Gamma ^ 2) +
            sum((delta - delta_new) ^ 2) / sum(delta ^ 2)
        if (control$trace) {
            cat("\nIteration ", iter, ":\n", sep = "")
            print(est)
            cat(rep("-", 70), "\n", sep = "")
        }
        if (tol < control$epsilon) return(est)
        lambda <- lambda_new
        Gamma <- Gamma_new
        delta <- delta_new
    }
    cat("Not converge yet after", iter, "iterations.\n")
    invisible(est)
}

### internal function
do1step <- function (lambda, Gamma, delta, m, n, x) {
    ## E-step
    logPmat <- outer(x, lambda, dpois, log = TRUE)
    fooList <- pois.HMM.lalphabeta(x, m, lambda, Gamma, delta)
    la <- fooList$la
    lb <- fooList$lb
    const <- max(la[, n])
    logL_T <- const + log(sum(exp(la[, n] - const)))
    lmu <- la + lb - logL_T
    lv1 <- array(NA, c(n - 1, m, m))
    lb <- t(lb)
    la <- t(la)
    ind <- 2 : n
    for (j in seq_len(m)) {
        lv1[, j, ] <- la[ind - 1, j] + logPmat[ind, ] + lb[ind, ]
    }
    ## M-step
    muMat <- exp(lmu)
    delta <- muMat[, 1]
    delta <- delta / sum(delta)
    f_jk <- Gamma / exp(logL_T) * colSums(exp(lv1), dims = 1)
    Gamma <- f_jk / rowSums(f_jk)
    lambda <- rowSums(muMat * rep(x, each = m)) / rowSums(muMat)
    list(lambda = lambda, Gamma = Gamma, delta = delta)
}

```

# Application to Earthquake Data

## Two-state HMM model

We first consider a two-state HMM model for the earthquake data.  The
sample mean of the data is about `r (xbar <- round(mean(qk7freq)))`.
Thus, we set the initial parameter of the Poisson distribution to be
`r paste(xbar + c(-1, 19), collapse = ", ")`, respectively.
For the transition probability matrix $\bm{\Gamma}$, the initial value
is a two-by-two matrix with off-diagonal elements 0.1.
The initial probability of each state is set to be equal, which is also
the default value of the implementation here.  The estimates are returned
if the algorithm converges.  The results are shown as follows:

```{r state2, cache = TRUE}
(lambda2 <- round(mean(qk7freq)) + c(- 1, 19))
Gamma2 <- matrix(c(0.9, 0.1, 0.1, 0.9), 2)
poisson.hmm.em(qk7freq, m = 2, lambda2, Gamma2)
```

## Three-state HMM model

We further consider a three-state HMM model. The initial parameter of
the Poisson distribution to be `r paste(xbar + c(-1, 9, 19), collapse = ", ")`,
respectively.  The initial value of the transition probability matrix is a
three-by-three matrix with off-diagonal elements 0.1. The initial probability
of each state is set to be equal as well.

```{r state3_25, cache = TRUE}
(lambda3 <- round(mean(qk7freq)) + c(- 1, 9, 19))
Gamma3 <- diag(0.8, 3)
Gamma3[upper.tri(Gamma3)] <- Gamma3[lower.tri(Gamma3)] <- 0.1
poisson.hmm.em(qk7freq, m = 3, lambda3, Gamma3)
```

By default, the maximum iteration times is 25.  Usually, we need more iterations
for the EM algorithm to converge.  We specify `maxit = 1000` in argument
`control` to allow at most 1000 iterations before convergence. The algorithm
converges this time and the estiamtes are returned as follows:

```{r state3_100, cache = TRUE}
poisson.hmm.em(qk7freq, m = 3, lambda3, Gamma3, control = list(maxit = 1e3))
```

## Trace the iterations

In addition, we may specify `trace = TRUE` in arugment `control` to trace each
iteration towards convergence. For example, the iterations of the two-state are
shown as follows:

```{r trace, cache = TRUE}
est <- poisson.hmm.em(qk7freq, m = 2, lambda2, Gamma2,
                      control = list(maxit = 5, epsilon = 1e-3, trace = TRUE))
```

Notice that the algorithm does not converge yet after only five iterations.

# Reference
