---
title: "STAT-6494 Advanced Statistical Computing with R"
subtitle: "Homework 5"
author: Wenjie Wang
date: "`r format(Sys.time(), '%d %B %Y')`"
header-includes:
    - \usepackage{booktabs}
    - \usepackage{bm}
bibliography: hw5.bib
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
    number_sections: yes
    toc: no
  fontsize: 11pt
  classoption: letter
  documentclass: article

---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
## function ipak: install and load multiple R packages.
## check to see if packages are installed.
## install them if they are not, then attach them to the search path.
ipak <- function (pkg) {
    new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
    if (length(new.pkg))
        install.packages(new.pkg, dependencies = TRUE)
    sapply(pkg, require, character.only = TRUE, quietly = TRUE)
    invisible(NULL)
}

## install packages needed
needpack <- c("knitr", "geepack")
ipak(needpack)
knitr::opts_chunk$set(dev = 'pdf')
knitr::opts_chunk$set(comment = NA)
## load data and clean

## source code

```


# Exercise: A Simple Implementation of QIF.

Write a function `simpleQIF` that evaluates the QIF $Q_n(\beta)$ at
regression coefficient vector $\beta$ for a given response vector $Y$,
covariate matrix $X$, an id vector that distinguishes the clusters,
and an exchangeable given working correlation structure.
For the example dataset `ohio`
in package **geepack**, feed your function to an optimizer such as `optim`
to solve for the parameter estimate.
An example call would look like `simpleQIF(y, X, id, family)`.

## Function `simpleQIF`

@qu2000improving proposed qudratic inference function (QIF) to
improve generalized estimating equations. Here, we consider
logit as link function for binary outcomes as default
and only assume exchangeable correlation structure.
The simple implementation of QIF, `simpleQIF` is designed to
return estimates of covariate coefficients.
The function body is given as follows.

```{r fun}
simpleQIF <- function (y, X, id, family = binomial) {
    X <- as.matrix(X)
    y <- as.vector(y)
    famFun <- family()
    ## balanced cluster size
    idTab <- as.numeric(table(id))
    clusN <- length(idTab)
    clusSize <- unique(idTab)
    if (length(clusSize) != 1) stop("Not balanced cluster size.")
    ## define matrix m_0 and matrix m_1
    ## for exchangeable structure
    mat0 <- diag(clusSize)
    mat1 <- matrix(1, clusSize, clusSize) - mat0
    matMList <- list(mat0, mat1)
    ## initialize beta as starting point
    fit0 <- glm.fit(X, y, family = famFun)
    beta0 <- fit0$coef
    ## optim step
    res <- nlm(qNbeta, beta0, clusSize = clusSize, clusN = clusN,
               y = y, X = X, matMList = matMList, famFun = famFun,
               stepmax = 10)
    ## simply output coef estimates
    setNames(res$estimate, colnames(X))
}

### internal functions
## computes g_ri(beta) for cluster i and matrix r
g_riBeta <- function (matM, betaVec, yi, Xi, famFun) {
    eta <- as.vector(Xi %*% betaVec, mode = "numeric")
    muiVec <- famFun$linkinv(eta)
    matDi <- famFun$mu.eta(eta) * Xi # element-wise
    sqrtInvAi <- diag(1 / sqrt(famFun$variance(muiVec)))
    crossprod(matDi, sqrtInvAi) %*% (matM %*% (sqrtInvAi %*% (yi - muiVec)))
}

## computes g_i(beta) for cluster i
g_iBeta <- function (betaVec, matMList, yi, Xi, famFun) {
    out <- lapply(matMList, g_riBeta, betaVec = betaVec, yi = yi,
                  Xi = Xi, famFun = famFun)
    do.call("c", out)
}

## computes QIF: Q_N(beta)
qNbeta <- function (betaVec, clusSize, clusN, y, X, matMList, famFun) {
    nMat <- length(matMList)
    nBeta <- length(betaVec)
    gNbeta <- matrix(NA, nrow = clusN, ncol = nMat * nBeta)
    indBase <- seq(1 - clusSize, 0)
    for (i in seq_len(clusN)) {
        ind <- indBase + i * clusSize
        Xi <- X[ind, ]
        yi <- y[ind]
        gNbeta[i, ] <- g_iBeta(betaVec, matMList, yi, Xi, famFun)
    }
    gBarN <- colMeans(gNbeta)
    cNbeta <- var(gNbeta)
    clusN * crossprod(gBarN, MASS::ginv(cNbeta) %*% gBarN)
}
```

## Comparison with **geepack** on a Sample Dataset

We compare the estimates of covariate coefficients produced by
`geese` in **geepack** to verify our implementation of QIF
on a sample dataset, *ohio*.

From the comparison, we may find that the estimates 
by `simpleQIF` and `geese` are fairly close to each other.

```{r comp, cache = TRUE}
require(geepack)
data(ohio)
y <- ohio$resp
X <- cbind("(Intercept)" = 1, ohio[, 3:4])
id <- ohio$id
simpleQIF(y, X, id) # estimates from our implementation
## compare with geepack
fitGee <- geese(resp ~ age + smoke, id = id, data = ohio, 
                family = binomial, corstr = "exch", scale.fix = TRUE)
fitGee$beta # estimates from geepack
```


# Reference
